SPP-Perc: Peceptron Learning Based Enhanced Signature Path Prefetcher

### 2018-12-07 design_arch.tex
done with pass

It can be beneficial to allow a prefetcher to speculate as deeply as
possible. Often, some useful prefetches are generated long after the
confidence of the prefetcher has fallen below a point, after which the
performance degrades, due to the increase of inaccurate prefetches. However,
in order to allow deep speculation in the prefetcher, inaccurate prefetches
must be filtered out. We propose to leverage perceptron-based learning as a
mechanism to differentiate between potentially useful deeply speculated
prefetches and likely not-useful ones. The Perceptron Prefetch Filter (PPF) is
placed between the prefetcher and the prefetch insertion queue, as illustrated
in Figure~\ref{fig:PPF_Hierarchy}, to prevent not-useful prefetches from
polluting the higher levels of the memory hierarchy.

Perceptron learning is a light-weight mechanism to pull together disparate
forms of information and synthesize a decision from them. Our work considers
a number of features corresponding to a prefetch, such as speculation depth,
page address and offset, and uses this information as the inputs to our
perceptron-based filter, in order to predict the usefulness of a prefetch. In
the remainder of this section, we discuss our proposed technique: PPF. It
enhances a base prefetcher, by filtering out not-useful prefetches. PPF is a
generalized prefetch filtering mechanism that may be adapted to any prefetcher
with appropriate modifications.

@@

Figure~\ref{fig:PPF_Datapath} shows the required structures of PPF, as well as
the steps followed to filter out not-useful prefetches. The filter is
organized as a set of tables, where each entry in the tables holds a
\textit{weight}. For a configuration of PPF using $N$ number of features, $N$
different tables of weights are needed. Each feature is used to index a
distinct table. The number of entries of each table varies according to their
corresponding feature, hence, different number of bits are needed to index
different tables. Each weight is a 5-bit saturating counter ranging from -16
to +15. We found that having 5-bit weights provide a good trade-off between
accuracy and area. A detailed explanation of the storage overhead of PPF can
be found in Section~\ref{Method-Overheads}.\\

\noindent{\textbf{Inferencing}\\
%ETG:is this like an explanation of how prefetching happens? 
%EB: I see this as a short context of the datapath leading to the perc filter.
The base prefetcher is triggered on every demand access to the L2 Cache. At
this point, it has the opportunity to trigger a prefetch. If it decides to do
so, it will also need to decide how many cache blocks to prefetch. These
blocks can be either placed in the L2 or L3 cache according to the confidence
of the prefetching mechanism. Once the base prefetcher is triggered, the
suggested prefetch candidates are fed to the perceptron filter to determine
the usefulness of these prefetches. The filter makes the decision to issue
the prefetch suggestions from the base prefetcher. To make the decision, each
feature corresponding to a suggested prefetch is used to index a table and all
the corresponding weights are summed. The sum denotes the confidence value
for the suggested prefetch, and is thresholded against two different
magnitudes: $\tau_{hi}$ and $\tau_{lo}$.

Prefetches whose sum exceeds $\tau_{hi}$ are placed into the L2 cache.
The higher confidence value hints the prefetch would be useful and should be
prioritized. A prefetch for which the features result in a confidence value
between $\tau_{lo}$ and $\tau_{hi}$ is allocated in the larger
LLC, as the filter is moderately confident of the future reuse of the cache
block, but not enough to possible pollute a significantly smaller L2.
Suggested prefetches for which the features lead to a confidence value lower
than $\tau_{lo}$ are not prefetched, as the low confidence value
represents that the perceptron has learn that similar set of features are
associated with non-useful prefetches. \newline

@@

In addition to the prefetch table mentioned above, PPF also maintains a
1,024-entry direct-mapped ``Reject Table.'' If a prefetch suggestion is
rejected by the perceptron layer, it is logged into the reject table. The
table is used to train the perceptron to avoid false negatives \textit{i.e.},
cases where the prediction suggested to reject the prefetch but the prefetch
was proven to be useful based on the observed demand accesses to the L2. \\

\noindent{\textbf{Feedback and Data Retrieval}\\
When there is an eviction or a demand access to the L2, training for both the
base prefetcher and our filter mechanism is triggered. The address of the
cache block that triggered training is used to index both the Prefetch and
Reject tables. 10 bits of the address are used to index tables, and another 6
bits to perform tag matching. Once the base prefetcher has suggested a
prefetch request, the corresponding features are used to index the tables of
weights. \newline

@@

\noindent{\textbf{Training}\newline
The address from the demand request triggering the training is looked up in
both tables. If the address is in the prefetch table and marked as valid, this
hints the previous prediction was correct and this is a useful prefetch. We
compute the sum of the corresponding weights. If the sum falls below a
specific threshold, training occurs and the corresponding weights are adjusted
accordingly. These thresholds are introduced in order to avoid over-training,
helping the filter adapt quickly to changes in memory behavior. These
thresholds are referred to as $\theta_p$ and $\theta_n$, 
respectively for the positive and negative values of training saturation.

@@

Prior work in this domain lacks the knowledge of which cache blocks should have 
been prefetched, but were not. The implementation of the reject table,
allows us to have this information, which can be use to further optimize
our prefetching mechanism.

@@

\subsection{Optimizing PPF for a Given Prefetcher}
\label{Arch-Generalizing}
The above discussion of PPF shows that it is highly modular and can be adapted
to be used over any base prefetcher for increased prefetch accuracy. Only 
two hooks are required between PPF and the baseline prefetcher.
The first is to make sure that all the prefetch candidates of the prefetcher
pass through the perceptron filter and if qualified, the metadata for
perceptron indexing be stored. The second is needed when the feedback of a
prior prefetch is available in form of a subsequent demand hit or cache
eviction. In that case, the stored metadata needs to be retrieved to update
the state of the perceptrons.

In general, PPF can be adapted to a new base prefetcher with only a
few modifications:\\
%
\textbf{Enhancing the Base Prefetcher:} By tuning down any internal thresholds
to increase its inherent aggressiveness. \\
%
\textbf{Inferencing and Storing:} All prefetch recommendations are tested
using the perceptron inferencing algorithm. The perceptron's output,
\textit{true} or \textit{false}, should be saved appropriately, along with all
metadata required for perceptron indexing. \\
%
\textbf{Retrieving and Training:} When feedback for a prefetch is available,
the previously stored metadata can be used to re-index into the perceptron
entries and increment or decrement the weights. \\
%
\textbf{Feature Selection:} Perceptrons essentially normalize contributions
from different features to get a single sum representing the final confidence.
So the perceptron learning can be as good as the set of features chosen.
Interestingly, this is what makes perceptron learning scalable, as it can
easily learn to incorporate newer information in the form of new features.
Some of the features we developed use information derived directly from
program execution, agnostic to the baseline prefetcher. Beyond that, the
feature set can be expanded to convey any useful information from the
prefetcher to the perceptron filter.

>>>>

It can be beneficial to allow a prefetcher to speculate as deeply as
possible. Often, some useful prefetches are generated long after the
confidence of the prefetcher has fallen below the point at which
performance degrades due to the increase of inaccurate prefetches.  In
order to allow deep speculation in the prefetcher, however, inaccurate
prefetches must be filtered out.  We propose to leverage
perceptron-based learning as a mechanism to differentiate between
potentially useful deeply speculated prefetches and likely not-useful
ones. The Perceptron Prefetch Filter (PPF) is placed between the
prefetcher and the prefetch insertion queue, as illustrated in
Figure~\ref{fig:PPF_Hierarchy}, to prevent not-useful prefetches from
polluting the higher levels of the memory hierarchy.

Perceptron learning is a light-weight mechanism to pull together
disparate forms of information and synthesize a decision from
them. Our work considers a number of features corresponding to a
prefetch, such as speculation depth, page address and offset, and uses
this information as the inputs to our perceptron-based filter in order
to predict the usefulness of a prefetch.  Here, we discuss our the
design of our proposed perceptron prefetch filter (PPF).  PPF enhances
a base prefetcher by filtering out predicted unused prefetches. PPF is
a generalized prefetch filtering mechanism that may be adapted to any
prefetcher with appropriate feature selection and modifications which
we describe below.

@@

Figure~\ref{fig:PPF_Datapath} shows the microarchitecture of PPF, as
well as the steps required to filter out not-useful prefetches. The
perceptron filter is organized as a set of tables, where each entry in
the tables holds a \textit{weight}.  For a configuration of PPF using
$N$ number of features, $N$ different tables of weights are needed.
Each feature is used to index a distinct table. The number of entries
of each table varies according to the corresponding feature, hence,
different number of bits are needed to index different tables. Each
weight is a 5-bit saturating counter ranging from -16 to +15. We found
that having 5-bit weights provides a good trade-off between accuracy
and area. A detailed explanation of the storage overhead of PPF can
be found in Section~\ref{Method-Overheads}.

\pg{the text below would be easier to follow if you used the figure
  more. make reference to the figure at the points that you are
  talking about}

\noindent \textbf{Inferencing}\\
%ETG:is this like an explanation of how prefetching happens? 
%EB: I see this as a short context of the datapath leading to the perc filter.
The base prefetcher is triggered on every demand access to the L2
Cache.  At this point, it has the opportunity to trigger a prefetch.
If it does so, it will also need to decide how many cache blocks to
prefetch.  These blocks can be either placed in the L2 or L3 cache
according to the confidence of the prefetching mechanism.  Once the
base prefetcher is triggered, the suggested prefetch candidates are
fed to the perceptron filter to determine the usefulness of these
prefetches. The filter ultimately decides whether to issue the
prefetch suggestions of the base prefetcher. To make the decision,
each feature corresponding to a suggested prefetch is used to index a
table and all the corresponding weights are summed. The sum denotes
the confidence value for the suggested prefetch, and is thresholded
against two different magnitudes: $\tau_{hi}$ and $\tau_{lo}$.

Prefetches whose sum exceeds $\tau_{hi}$ are placed into the L2 cache.
The higher confidence value hints the prefetch would be useful and
should be prioritized.  A prefetch for which the features result in a
confidence value between $\tau_{lo}$ and $\tau_{hi}$ is allocated in
the larger LLC, as the filter is moderately confident of the future
reuse of the cache block, but not enough to possible pollute a
significantly smaller L2.  Suggested prefetches for which the features
lead to a confidence value lower than $\tau_{lo}$ are not prefetched,
as the low confidence value represents that the perceptron learned
that a similar set of features are associated with non-useful
prefetches.

@@

In addition to the prefetch table mentioned above, PPF also maintains
a 1,024-entry direct-mapped ``Reject Table.'' If a prefetch suggestion
is rejected by the perceptron layer, it is logged into the reject
table. The table is used to train the perceptron to avoid false
negatives \textit{i.e.}, cases where the prediction suggested to
reject the prefetch but the prefetch was proven to be useful based on
the observed demand accesses to the L2.

\noindent \textbf{Feedback and Data Retrieval}\\
When there is an eviction or a demand access to the L2, training for
both the base prefetcher and our filter mechanism is triggered. The
address of the cache block that triggered training is used to index
both the Prefetch and Reject tables. 10 bits of the address are used
to index tables, and another 6 bits to perform tag matching. Once the
base prefetcher has suggested a prefetch request, the corresponding
features are used to index the tables of weights.

@@

\noindent \textbf{Training}\newline The address from the demand
request triggering the training is looked up in both tables. If the
address is in the prefetch table and marked as valid, this hints the
previous prediction was correct and this is a useful prefetch. We
compute the sum of the corresponding weights. If the sum falls below a
specific threshold, training occurs and the corresponding weights are
adjusted accordingly. These thresholds are introduced in order to
avoid over-training, helping the filter adapt quickly to changes in
memory behavior. These thresholds are referred to as $\theta_p$ and
$\theta_n$, respectively for the positive and negative values of
training saturation.

@@

\pg{what prior work? this reference is too oblique to be useful,
  either be specific or drop this paragraph} Prior work in this domain
lacks the knowledge of which cache blocks should have been prefetched,
but were not. The implementation of the reject table, allows us to
have this information, which can be use to further optimize our
prefetching mechanism.

@@

\subsection{Optimizing PPF for a Given Prefetcher}
\label{Arch-Generalizing}
The above discussion of PPF shows that it is highly modular and can be
adapted to be used over any base prefetcher for increased prefetch
accuracy.  Only two hooks are required between PPF and the baseline
prefetcher.  The first is to make sure that all the prefetch
candidates of the prefetcher pass through the perceptron filter and if
qualified, the metadata for perceptron indexing be stored. The second
is needed when the feedback of a prior prefetch is available in form
of a subsequent demand hit or cache eviction. In that case, the stored
metadata needs to be retrieved to update the state of the perceptrons.
\pg{this isn't quite right, I thought we were taking some metadata
  from the prefetcher itself like confidence.  You should explicitly
  mention that prefetcher metadata can be used, update the figure to
  show that metadata passing from SPP to the prefetcher and point to
  the figure to show where that connection is}

In general, PPF can be adapted to a new base prefetcher with only a
few modifications:

\noindent \textbf{Make Base Prefetcher More Aggressive:} Tune down any
internal thresholds or throttling mechanisms to increase its
aggressiveness.

\noindent \textbf{Inferencing and Storing:} All prefetch
recommendations are tested using the perceptron inferencing algorithm.
The perceptron's output, \textit{true} or \textit{false}, should be
saved appropriately, along with all metadata required for perceptron
indexing.

\noindent \textbf{Retrieving and Training:} When feedback for a
prefetch becomes available, the previously stored metadata can be used
to re-index into the perceptron entries and increment or decrement the
weights.

\noindent \textbf{Feature Selection:} Perceptrons essentially
integrate contributions from different features to get a single sum
representing the final confidence.  Thus, perceptron learning can only
be as good as the set of features chosen.  Interestingly, this is what
makes perceptron learning scalable, as it can easily learn to
incorporate newer information in the form of new features.  Some of
the features we developed use information derived directly from
program execution, agnostic to the baseline prefetcher.  Beyond that,
the feature set can be expanded to convey any useful information or
metadata available in the baseline prefetcher.
